---
title: "Lecture 8"
subtitle: "Principles of Honest Charts & Visualizing Quantities & Distributions"
author: Saryu Onishi
date: HES 3000-002, Sport Performance Analytics
output:
  xaringan::moon_reader:
    seal: false
    css: ['default', 'metropolis', 'metropolis-fonts', 'ou-colors.css']
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'

---

class: title-slide    

<br><br><br>

# Lecture 8

## Principles of Honest Charts & Visualizing Quantities & Distributions

### Saryu Onishi

### HES 3000-002, Sport Performance Analytics

---

# Week 5, Class 1 Overview

- **Where We Are:**
  - Last week, we mastered descriptive statistics (center, spread, distributions) and ran preliminary stats on your group project data.
  
- **Today's Focus:**
  - Transitioning from numbers to pictures: The principles of **honest charts** and how to visualize **Quantities and Distributions** effectively.

- **The Big Goal:**
  - Learn how to build clear, truthful visualizations in R that communicate insights to coaches and staff without misleading the audience.

---

# Why Data Visualization Matters in Sports

- **The Human Brain and Visual Processing:**
  - Coaches and General Managers rarely have time to parse dense rows of numbers in an Excel sheet. 
  - A clean, well-designed chart communicates complex findings in seconds.

- **The Analyst's Responsibility:**
  - Visuals are persuasive. If your chart is confusing or misleading, you will misinform key decision-makers and lose organizational trust.

---

# The Power (and Danger) of Visuals

- **The Double-Edged Sword:**
  - Charts can illuminate hidden performance trends... or they can completely distort reality to fit a pre-existing narrative.

- **Bad Visualization in Media:**
  - Sports media is notorious for sensationalized graphics: truncated axes, missing baselines, and misleading 3D bar charts that exaggerate minor performance differences.

---

# Core Principles of Honest Charts

- **1. Show the Data Clearly:**
  - Never let design fluff (unnecessary 3D effects, busy backgrounds, excessive gridlines) obscure the actual data points.

- **2. Do Not Misrepresent Magnitude:**
  - Visual elements must scale directly and proportionally to the numerical values they represent.

- **3. Provide Complete Context:**
  - Always include clear labels, units of measurement, and appropriate baselines.

---

# Principle 1: Scale Integrity (The Axis Trap)

- **The Truncated Y-Axis:**
  - Starting a bar chart at a non-zero baseline artificially exaggerates small differences between players or teams.
  
- **Sports Example:**
  - If Team A averages 102 points per game and Team B averages 105, starting the y-axis at 100 makes Team B look twice as good as Team A, when the actual difference is negligible.

- **Rule of Thumb:**
  - For bar charts representing counts or totals, the y-axis **must** start at zero.

---

# Principle 2: Proportional Area & Visual Weight

- **The 2D vs. 1D Trap:**
  - When scaling circles or 3D objects, human eyes judge **area or volume**, not just linear radius.
  - Doubling the radius of a circle quadruples its area, creating a massive visual distortion of value.

- **The Fix:**
  - Stick to standard 2D Cartesian coordinates (bars, dots, lines) where length and position map cleanly to numerical values.

---

# Principle 3: Context and Comparison

- **Numbers Need Denominators:**
  - Visualizing raw totals (e.g., total passing yards) often just highlights who played the most minutes.
  - Honest charts visualize **rates, efficiencies, or per-minute metrics** to provide fair comparisons across athletes.

---

# Visualizing Quantities in Sports

- **What is a Quantity Chart?**
  - Comparing discrete values across categories (e.g., goals by player, payroll by team, distance by position).

- **Primary Tools:**
  - **Bar Charts / Column Charts:** The gold standard for comparing quantities.
  - **Lollipop Charts:** A cleaner, modern alternative to crowded bar charts.
  - **Cleveland Dot Plots:** Excellent for comparing multiple metrics across categories without visual clutter.

---

# Visualizing Quantities Best Practices

- **Ordering Matters:**
  - Never leave categorical bars in alphabetical order unless it's a true alphabetical scale. 
  - **Always sort bars by value** (descending or ascending) so the coach's eye instantly sees who is first and last.

- **Horizontal vs. Vertical:**
  - Use horizontal bar charts when category labels (player names, team names) are long to prevent unreadable text warping.

---

# Visualizing Distributions in Sports

- **Why Look at Distributions?**
  - Means and medians hide variance. Visualizing distributions shows you the *full shape* of an athlete's performance range.

- **Primary Tools:**
  - **Histograms:** Showing frequency bins for continuous data.
  - **Density Plots:** Smoothed representations of data distributions.
  - **Boxplots:** Summarizing the median, IQR, and outliers compactly.
  - **Violin Plots:** Combining density curves with boxplots for deep distributional insights.

---

# Visualizing Distributions: Boxplots in Sports

- **Anatomy of a Boxplot:**
  - **The Center Line:** Median.
  - **The Box:** Interquartile Range (middle 50% of performances).
  - **The Whiskers:** Spread of typical data (up to 1.5 * IQR).
  - **Points Beyond Whiskers:** Outliers (exceptional breakout games or terrible slumps).

- **Sports Application:**
  - Comparing the consistency of two players who both average 20 points per game, but have drastically different boxplot spreads.

---

# Case Study: Spotting Skewness Visually

- **The Scenario:**
  - You analyze player salaries or rare event metrics (like turnovers or fast-break points).
- **What a Histogram Reveals:**
  - A heavily right-skewed distribution where 90% of players cluster at the low end, and a few superstars stretch the tail far to the right.
- **Why It Matters:**
  - Visualizing the skew tells you immediately that using the mean will mislead management—you *must* use the median and IQR instead.

---

# Introduction to ggplot2 in R

- **The Grammar of Graphics:**
  - Built on the philosophy that charts are built layer by layer: Data -> Aesthetic Mappings -> Geometric Objects (geoms).

- **Basic Syntax Structure:**
```{r, eval = FALSE}
library(ggplot2)

ggplot(data = df, aes(x = variable1, y = variable2)) +
  geom_point() +
  theme_minimal()
```

---

# Code Example: Visualizing Quantities in R

- Creating a Clean Bar Chart:
```{r, eval = FALSE}
library(ggplot2)
library(dplyr)

# Summarize and plot mean metric by position
df_summary <- df %>%
  group_by(position) %>%
  summarize(mean_val = mean(metric_name, na.rm = TRUE))

ggplot(df_summary, aes(x = reorder(position, mean_val), y = mean_val)) +
  geom_col(fill = "steelblue") +
  coord_flip() + # Horizontal for readability
  labs(title = "Mean Performance by Position", x = "Position", y = "Value") +
  theme_minimal()
```

---

# Code Example: Visualizing Distributions in R

- Creating a Boxplot by Group:
```{r, eval = FALSE}
ggplot(df, aes(x = position, y = metric_name, fill = position)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Distribution of Performance Metric by Position", 
       x = "Position", y = "Metric Value") +
  theme_minimal() +
  theme(legend.position = "none")
```

---

# In-Class Lab: Building Honest Charts (20 mins)

- Lab Objectives:
 - Open your group project R script.
 - Build at least one quantity chart (e.g., bar chart of means by position/player) and one distribution chart (e.g., boxplot or histogram) using ggplot2.
 - Apply the principles of honest charts: clean labels, zero-baselines where appropriate, and sorted categories.

- Deliverable:
 - Two functional ggplot code chunks saved in your project repository script.

---

# Lab Share-Out & Peer Critique

- Discussion:
 - Let's project 1 or 2 group charts on the screen.
 - Do these charts follow our honesty principles? Are the axes clear? Does the visualization tell an immediate story?

- Instructor Feedback:
 - Refining your visual communication now will make your final project presentation drastically stronger at the end of the semester.

---

# Summary & Looking Ahead

- Key Takeaway: Honest charts protect your credibility. Use quantity charts to compare categories fairly and distribution charts to reveal the full story behind the averages.

- Next Class (Week 5, Class 2): We cover Visualizing Proportions and Relationships, followed by a hands-on project lab where you apply these visualization techniques directly to your project datasets.

- For Next Time: Ensure your basic quantity and distribution plots are generated and saved in your project repo.