---
title: "Understanding Your Data: Characteristics & Quality"
subtitle: "HES 3000: Sport Performance Analytics"
author: "Lecture Presentation"
date: "Fall 2026"
output: 
  ioslides_presentation:
    widescreen: true
    smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
library(dplyr)
library(ggplot2)
```

## The Data Hunt: Recap

**You found data. Now what?**

In our last activity, you hunted down sports performance datasets. You might have downloaded a CSV of game stats, pulled an open-source tracking dataset, or exported wellness questionnaires.

Before we can analyze, visualize, or model this data, we have to understand its "personality."

**Today's Goals:**
1. Identify the fundamental types and structures of your data.
2. Understand sampling rates and time-series nuances in sports tech.
3. Assess the "shape" (Wide vs. Long) of your data.
4. Evaluate data quality (Validity, Reliability, Missingness).

---

## The Anatomy of a Dataset

In most applied analytics, we look at rectangular data (data frames).

- **Rows = Observations:** Depending on the context, a single row could be:
  - One athlete's daily session summary.
  - One jump on a force plate.
  - One millisecond of GPS tracking data.
- **Columns = Variables:** The specific metrics or identifiers being recorded (e.g., `Athlete_ID`, `Session_Date`, `PlayerLoad`, `Max_Velocity`).

*Question for your dataset:* What exactly does one single row represent? If you don't know this, you cannot analyze the data.

---

## Data Types in R: What are we looking at?

When you load your data into R, the software categorizes every column. Misunderstood data types lead to broken code.

*   **Numeric (Continuous):** Can be any value, including decimals.
    *   *Example:* Sprint speed (32.4 km/h), Peak Force (2450.5 N).
*   **Integer (Discrete):** Whole numbers. 
    *   *Example:* Number of sprints > 20km/h, Goals scored.
*   **Character / String:** Text data.
    *   *Example:* "Forward", "Home", "Away".
*   **Factor:** Categorical data with a set number of levels.
    *   *Example:* RPE (Rate of Perceived Exertion) from 1 to 10.
*   **Date / POSIXct:** Time data (crucial for sports!).
    *   *Example:* `2026-09-09 14:30:00`

---

## Code Check: Getting to Know Your Data

The first step after importing your CSV is looking under the hood. The `glimpse()` function from `dplyr` is your best friend.

```{r}
# Assuming you loaded a dataset called 'gps_session'
library(dplyr)

glimpse(gps_session)
```

**What to look for in the output:**
- Are your dates actually reading as `<date>` or did R read them as `<chr>` (text)?
- Are your metrics (like `PlayerLoad`) reading as `<dbl>` (numeric), or did a stray comma turn them into characters?

---

## Sensor Nuance: Sampling Rates (Hz)

In sport performance, the *frequency* of data collection fundamentally changes how you handle it.

**Low Frequency (Daily/Weekly)**
- *Example:* Wellness questionnaires, 1RM testing.
- *Volume:* Small. Easy to view in Excel.

**Medium Frequency (10Hz - 100Hz)**
- *Example:* Catapult Vector or other micro-sensor units. (10 data points per second).
- *Volume:* Large. A 90-minute session generates 54,000 rows *per athlete*.

**High Frequency (1000Hz+)**
- *Example:* VALD ForceDecks. (1000 data points per second).
- *Volume:* Massive. A 3-second CMJ generates 3,000 rows of raw data for the left leg, right leg, and total.

---

## Data Structure: Wide vs. Long Data

The shape of your data dictates how easily you can visualize it in `ggplot2`.

**Wide Format (Human Readable)**
Great for viewing, terrible for graphing. Every variable gets its own column.
*   *Athlete | Match_1_Load | Match_2_Load | Match_3_Load*

**Long Format (Machine Readable - Tidy Data)**
Crucial for `ggplot2`. Variables are collapsed into key-value pairs.
*   *Athlete | Match_Number | Load*
*   *Smith | 1 | 450*
*   *Smith | 2 | 480*

---

## Data Quality: The Unsexy Reality

*"Garbage in, garbage out."* 

Before trusting your data, ask these three questions:

1. **Validity:** Does this metric measure what it claims to measure? 
   - *Example:* Is "High Speed Distance" capturing true sprint effort, or just a slow jog for a fast athlete?
2. **Reliability:** If the same athlete does the exact same movement tomorrow, will the sensor output the exact same number?
3. **Missingness:** Why is there an `NA`? 
   - Did the athlete forget to fill out the survey?
   - Did the hardware battery die?
   - Was the athlete injured? (Missing data is sometimes the most important signal!).

---

## Activity: Interrogate Your Dataset (15 Minutes)

Open up RStudio, load the data you found in our last class, and answer these questions with a partner:

1. Use `glimpse()`. How many observations (rows) and variables (columns) do you have?
2. What does a single row represent in your dataset?
3. Identify one continuous variable, one categorical (factor) variable, and one character variable.
4. Do you see any missing values (`NA`)? If so, why do you think they are missing?

```{r}
# Starter code for the activity:
my_data <- read.csv("your_downloaded_data.csv")
glimpse(my_data)
summary(my_data)
```

---

## Looking Ahead: Data Cleaning

Now that you know what your data looks like, you probably noticed it's a bit messy. 

Next week, we start **Data Cleaning**. We will use `dplyr` to:
- Filter out bad rows.
- Select the columns we actually care about.
- Mutate (create) new performance variables.

**Homework:** Ensure your dataset is successfully loaded into RStudio and your workspace is saved for next class.
